Neural networks for solving regression and classification problems

### 2022-11-15 doc/paper.tex
Paper

From FIGURE MOMENTUM v CONSTANT we can observe that adding momentum reduces the amount of epochs necessary to reach a lower MSE score. Already after EPOCHS epochs, the momentum scheduler has achieved MOMENTUM MSE AT EPOCHS, a notable improvement over CONSTANT MSE AT EPOCHS achieved by a constant learning rate.

- GD vs SGD

Furthermore, with the introduction of various adaptive scheduler methods such as RMS prop, adagrad, adagrad with momentum and adam we can observe from FIGURE ALL that we can reach even greater results in fewer epochs. We can also observe that the different schedulers that they all reach REASONABLE MSE within a couple of epochs, but in order to get a better understanding of the computational cost we have also compared the time it took to fit the data. Looking at TIME TABLE FOR ALL SCHEDULERS, BEST took the least time, beating out our baseline constant scheduler by TIME DIFF. Surprisingly, RMS prop performed more poorly than expected.

An interesting insight from FIGURE NUM is that all the schedulers prefer high learning rates and regularization. As this plot was generated for 100 epochs due to the computational complexity of gridsearching the six schedulers, the bias towards higher learning rates is certainly due to the high learning rates being able to reach a low MSE in the short training time. With the lambda = 0 column representing the ordinary least squares solutions, these are clearly consistantly outperformed by Ridge regression as the best MSE is achieved with regularization. For this specific plot, momentum does not seem to be an important parameter. Both the momentum and adagrad with momentum heatmaps are identical to their momentumless counterparts, constant and adagrad, implying that the best gridsearched momentum is 0. That momentum is irrelevant in this instance is not a generalizable result, as momentum has played a key role in previous plots and results.

Another observation from FIG NUM  is that the schedulers final error for100 epochs are more or less the same. As the largest change in MSE occurs in the first 50 epochs as seen in ALL FIGS, perhaps the other schedulers are given enough time to catch up as the best performing schedulers, such as Adam, starts slowing down the rate at which the MSE is decreased. However, as can be seen in HEATMAP FIG, Adam offers a 11.11 \% \footnote{$\frac{constant MSE}{adam MSE} * 100 = \frac{0.018}{0.016} * 100 = 11.11\%$} decrease in MSE compared to the constant scheduler. It is clear from the results thus far that Adam is the best scheduler for approximating the Franke function when using linear regression with gradient descent.

Furthermore, when trained for 5000 epochs, Adam with optimal parameters keeps improving the MSE. Scikit learn's Adam scheduler does perform better, offering a greater decrease the MSE. Even though we have cross validated both our own Adam implementation and Scikit's with 5 folds, these differences in results might be due to a more favorable weight initiation in Scikits model, in addition to other optimizations present in the scikit learn implementation. Nonetheless, neither models are able to beat the analytical solution provided by linear regression methods using polynomial approximations, with our best result of an MSE of 0.012 on the test data and Scikits final MSE of 0.004 (granted this is on the train data, as scikit learn does not provide the loss curve for the test data).  both falling short of the analytical MSE of 0,003. In theory, the feed forward neural network should be able to better approximate the Franke function given that it is not limited to polynomial solutions. This result might be due to 5000 epochs not being enough training time, as the Franke function is a quite complex function with added noise.

@@

@@

To test our neural networks ability to solve classification problems, we first applied it to the Wisconsin Breast Cancer dataset. Using our gridsearch method, we found that the best scheduler was Adam, with an eta of $0.01$ and no regularization. For our architecture, we chose to use only one hidden layer with 60 nodes, based on our testing. With this, we were able to achieve a crossvalidated test accuracy of $98.6 \%$, which is very good. Looking at the confusion matrix (Fig.~\ref{cancerconf}), we see that we are achieving good predictions for both true positives and true negatives.

\begin{figure}[h]
\centerline{\includegraphics[width=5in]{figure/cancerconf.png}}
\caption{Confusion matrix for the cancer dataset using a neural net with a single hidden layer of 60 nodes}
\label{verification}
\end{figure}

Confident in our networks's ability to to binary classification, we turned to the MNIST dataset. Using our gridsearch, we again found that the best scheduler was Adam. Our architecture in this case was a single hidden layer with 100 nodes. Being a multiclass problem with ten different classes, one might expect our neural network to struggle more with this dataset. However, this was not the case, as we achieved a crossvalidated test accuracy of $99.07$. From these two datasets, we can see that a correctly optimized neural network is adept at solving classification problems. On the other hand, these good results, and the easew with which our network attained them, could imply that the problems we have dealt with are not as complicated as they may seem.

@@

Based on our good results with the neural net, we attempt to solve the same problems using simple logistic regression. Using logistic regression on the cancer dataset, we found through our gridsearch that Adam with an eta $0.1$ gave the best results. Here we found that logistic regression performed essentially equivalent or even better, at a much lower computational cost. Looking at Fig.~\ref{cancerlogneur}, we see that logistic regression keeps up and even surpasses our neural network in accuracy per epoch. The value of logistic regression becomes even greater when we consider the computational cost of each method. In this case, the logistic regression was about 65 \% faster\footnote{The time taken to complete a crossvalidation with 5 folds up to 200 epochs was about 52 seconds for our neural network, but only 18 seconds for logistic regression}, with results surpassing our neural network (see Fig.~\ref{cancerconflog}). This implies that our classification datasets were perhaps too simple, and that using Neural Networks was not needed to achieve good results.

>>>>

From FIGURE MOMENTUM v CONSTANT we can observe that adding momentum reduces the amount of epochs necessary to reach a lower MSE score. Already after EPOCHS epochs, the momentum scheduler has achieved MOMENTUM MSE AT EPOCHS, a notable improvement over CONSTANT MSE AT EPOCHS achieved by a constant learning rate. Furthermore, with the introduction of various adaptive scheduler methods such as RMS prop, adagrad, adagrad with momentum and adam we can observe from FIGURE ALL that we can reach even greater results in fewer epochs. We can also observe that the different schedulers that they all reach REASONABLE MSE within a couple of epochs, but in order to get a better understanding of the computational cost we have also compared the time it took to fit the data. Looking at TIME TABLE FOR ALL SCHEDULERS, BEST took the least time, beating out our baseline constant scheduler by TIME DIFF. Surprisingly, RMS prop performed more poorly than expected.

Before continuing, it is interesting to note that all the schedulers work best with a combination of similar eta and lambda values when gridsearched for achieving the minimal MSE. It is also interesting that they in the long term reach more or less the same error as can be observed in HEATMAP FIG. This could be because given more epochs, the other schedulers are given more time to catch up as the best performing schedulers, such as Adam, only decrease their MSE slightly. However, as can be seen in HEATMAP FIG, Adam offers a $\frac{0.018}{0.016} \dot 100 = 11.11\%$ decrease in MSE compared to the constant schedulers.

However, when trained for 500 epochs, Adam with optimal parameters does not yield any improvement in MSE. Scikits Adam scheduler does perform better, further decreasing the MSE slightly. Even though we have cross validated both our own Adam implementation and Scikit's with 5 folds, these differences in results might be due to a more favorable weight initiation in Scikits model. Nonetheless, neither models are able to beat the analytical solution provided by linear regression methods using polynomial approximations, with our best result of ADAM FINAL MSE and Scikits final MSE of SCIKIT FINAL MSE both falling short of the analytical MSE of 0.003. In theory, the feed forward neural network should be able to better approximate the Franke function given that it is not limited to polynomial solutions. This result might be due to 500 epochs not being enough training time.

@@

Having gained a great understanding of how the different schedulers function, and having created the necessary infrastructure to study our models performance based on a number of different factors, we use these methods to optimize our neural network for linear regression. More specifically, we attempt to fit the Franke function. First, we perform a preliminary test on MSE over model complexity. For comparisons sake, we use the same scheduler for each architecture, and fix the regularization parameter at 0. By fixing the regularization parameter at 0, we are of course favouring less complex models, but this a good way to get a general sense of when the model starts to overfit. Using the sigmoid activation function, we can study the MSE as a function of the total hidden node count (Fig.~\ref{msebymodelfranke}). Here we can see that we are not overfitting

\begin{figure}[h]
\centerline{\includegraphics[width=6in]{figure/msebymodelfranke.png}}
\caption{Doge}
\label{msebymodelfranke}
\end{figure}

@@

To test our neural networks ability to solve classification problems, we first applied it to the Wisconsin Breast Cancer dataset. But to get the best results, there are a large amount of parameters that need to be decided on. We must find the best architecture, the best, activation functions, best scheduler and the best parameters for the scheduler. Finding the absolute best combination of all these choices is impossible, so compromises must be made.

We start by trying to find a good model complexity, i.e the best number of hidden layers, and the amount of nodes in them. To do this, we plotted the accuracy over different model architectures. Like we did during regression, we fix all the other parameters and study the architecture by itself. Using the sigmoid activation function (Fig.~\ref{cancerarchcompsigmoid}), our results are somewhat surprising. We see that even at 800 total hidden nodes, we are not overfitting. Regardless of this peculiarity, we can see that the gain from using more layers and more nodes is miniscule. Using the RELU activation function, the results are quite different, as can be seen in Fig.~\ref{cancerarchcomp}. Here we can observe that the neural net with two hidden layers starts performing very inconsistently when the number of nodes increases. Looking more closely at the training process, we can see that training stops completely in some of our crossvalidations. Clearly, we are running into the "dying RELU problem", where too many of our RELU nodes output zero, leading to our gradient descent slowing down and eventually stopping completely. Here we can clearly see some of the benefits of the sigmoid, as even if the gradients become very small, they will never be zero, as with RELU. Based on these findings, we decided to use a simple model with a single hidden layer of 100 nodes.

\begin{figure}[h]
\centerline{\includegraphics[width=6in]{figure/cancerarchcompsigmoid.png}}
\caption{Comparison between a neural net with one hidden layer, and one with two hidden layers, for different totals of hidden nodes. For the network with two hidden layers, the total is evenly divided between the layers, such that at $x = 60$ we have a network with two hidden layers of 30 nodes each. In this case, the hidden layers used the sigmoid activation function.}
\label{cancerarchcompsigmoid}
\end{figure}

\begin{figure}[h]
\centerline{\includegraphics[width=6in]{figure/cancerarchcomp.png}}
\caption{Comparison between a neural net with one hidden layer, and one with two hidden layers, for different totals of hidden nodes. For the network with two hidden layers, the total is evenly divided between the layers, such that at $x = 60$ we have a network with two hidden layers of 30 nodes each. In this case, the hidden layers used the RELU activation function.}
\label{cancerarchcomp}
\end{figure}

Using our gridsearch method on this model, we found that the best scheduler was Adam, with an eta of $0.001$ and a lambda of $0.0001$. With this combination of architecture and scheduler, we were able to achieve a crossvalidated test accuracy of $98.76\%$, which is a great result. This outperforms SciKit's MLP classifier, which achieved a crossvalidated test accuracy of around 95\%. Looking at the confusion matrix (Fig.~\ref{cancerconf}), we see that we are achieving good predictions for both true positives and true negatives.

\begin{figure}[h]
\centerline{\includegraphics[width=5in]{figure/cancerconf.png}}
\caption{Confusion matrix for the cancer dataset using a neural net with a single hidden layer of 60 nodes}
\label{cancerconf}
\end{figure}

Confident in our networks's ability to do binary classification, we turned to the MNIST dataset. Here, we perform an educated guess on the different parameters, based on our previous testing. We decided upon an architecture of a single layer with 100 nodes using the RELU activation function. For our scheduler, we used Adam with an eta of $0.001$ and no regularization. Being a multiclass problem with ten different classes, one might expect our neural network to struggle more with this dataset. However, this was not the case, as we achieved a crossvalidated test accuracy of $99.07\%$. Based on these results, we conclude that the initial guess for our model was a good one, and that further optimization has little room for improvement. From these two datasets, we can see that a neural network is adept at solving classification problems. On the other hand, these good results, and the ease with which our network attained them, could imply that the problems we have dealt with are not as complicated as they may seem.

@@

Based on our good results with the neural net, we attempt to solve the same problems using simple logistic regression. Using logistic regression on the cancer dataset, we found through our gridsearch that Adam with an eta $0.1$ gave the best results. Here we found that logistic regression performed essentially equivalent or even better, at a much lower computational cost. Looking at Fig.~\ref{cancerlogneur}, we see that logistic regression keeps up and even surpasses our neural network in accuracy over epochs. The value of logistic regression becomes even greater when we consider the computational cost of each method. In this case, the logistic regression was about 65\% faster\footnote{The time taken to complete a crossvalidation with 5 folds up to 200 epochs was about 52 seconds for our neural network, but only 18 seconds for logistic regression}, with results surpassing our neural network (see Fig.~\ref{cancerconflog}). This implies that our classification datasets were perhaps too simple, and that using Neural Networks was not needed to achieve good results.
